### 2012-11-05 Docs/Thesis/abstract.tex
Thesis edits

All signal-function implementations of FRP to date have utilized demand-driven
or ``pull-based''  evaluation for both events and signals, producing output from
the FRP system whenever the consumer of the output is ready. This greatly
simplifies the implementation of signal-function FRP systems, but leads to
inefficient and wasteful evaluation of the FRP system when this strategy is
employed to evaluate events.

In contrast, an input-driven or ``push-based'' system evaluates the network
whenever new input is available. This frees the system from evaluating the
network when nothing has changed, and then only components necessary to react to
the input are re-evaluated. This form of evaluation has been applied to events
in standard FRP systems~\cite{Elliott2009} but not in signal-function FRP
systems.

I describe the design and implementation of a signal-function FRP system which
applies pull-based evaluation to signals and push-based evaluation to events (a
"push-pull" system). The semantics of the system are discussed, and its
performance and expressiveness for practical examples of interactive programs
are compared to existing signal-function FRP systems through the implementation
of a networking application.

>>>>

All signal-function implementations of FRP to date have utilized demand-driven
or ``pull-based''  evaluation for both events and signals, producing output from
the FRP system whenever the consumer of the output is ready. This greatly
simplifies the implementation of signal-function FRP systems, but leads to
inefficient and wasteful evaluation of the FRP system when this strategy is
employed to evaluate events, because the components of the signal function which
process events must be computed whether or not there is an event occurrence.

In contrast, an input-driven or ``push-based'' system evaluates the network
whenever new input is available. This frees the system from evaluating the
network when nothing has changed, and then only the components necessary to
react to the input are re-evaluated. This form of evaluation has been applied to
events in standard FRP systems~\cite{Elliott2009} but not in signal-function FRP
systems.

I describe the design and implementation of a signal-function FRP system which
applies pull-based evaluation to signals and push-based evaluation to events (a
``push-pull'' system). The semantics of the system are discussed, and its
performance and expressiveness for practical examples of interactive programs
are compared to existing signal-function FRP systems through the implementation
of a networking application.

### 2012-11-05 Docs/Thesis/appendix-types.tex
Thesis edits

\begin{description}
\item[ADT] See {\em Algebraic Datatype}

@@

\item[Existential Type] An existential type is a means of ``hiding'' a
{\em type variable} in an {\em Algebraic Datatype} or
{\em Generalized Algebraic Datatype}. The Glasgow Haskell Compiler provides
existential types in two ways. The first is the {\tt Rank2Types} and
{\tt RankNTypes} extensions. With either of these extensions enabled,
type variables may be brought into scope for specific {\em data constructor} in
an Algebraic Datatype using the {\tt forall} keyword. Since this
type variable is not contained in the parameters of the type constructor, it
cannot be constrained by type inference. Thus, no external function can assume
anything about its type, except for equality between occurrences of the
variable, and {\em class constraints} given in the definition of the data
constructor. The other way is through Generalized Algebraic Datatypes. Any
variable occurring in the type of a data member of a data constructor of a
Generalized Algebraic Datatype, but not occurring in the type of that
constructor, is automatically existentially quantified.

@@

@@

\item[Type Variable] A type variable is a type-level term which may be 
instantiated (by the typechecker via inference, or by the user via annotation)
with any type. Together with {\em type constructors}, type variables form the
basis of polymorphism in Haskell and similar languages.
\end{description}

>>>>

\begin{description}
\item[ADT] See {\em Algebraic Datatype}.

@@

\item[GADT] See {\em Generalized Algebraic Datatype}.

@@

\item[Kind] A ``type of types.'' Kinds are used to verify that types are
consistent during typechecking. The kind of types which contain values is
{\tt *}, and the kind of single-parameter type constructors which take a
type is {\tt * -> *}. Other kinds may also be introduced. For instance,
signal vectors should be their own separate kind, but the Haskell type mechanism
was not mature enough to support this at the time of this writing..

@@

\item[Type Variable] A type variable is a type-level term which may be 
instantiated (by the typechecker via inference, or by the user via annotation)
with any type at the point where the value so typed is used. 
Together with {\em type constructors}, type variables form the basis of
polymorphism in Haskell and similar languages.
\end{description}

### 2012-11-05 Docs/Thesis/background.tex
Thesis edits

The key abstraction of functional programming is the function, which takes an input and produces an output. Functions may produce functions
as output and/or take functions as input. We say that functions are {\em first-class values} in a language.

@@

The goal of FRP is to provide compositional and high-level abstractions for creating reactive programs. The key
abstractions of FRP are behaviors or signals\footnote{Behaviors are generally referred to as {\em signals} in signal function literature. This is unfortunate, since a signal
function may operate on events, signals, or some combination of the two.}, which are time-dependent values defined at every point in continuous time, and events, which are 
values defined at countably many points in time. An FRP system will provide combinators to manipulate events and signals and to react
to events by replacing a portion of the running program in response to an event. Behaviors and events, or some abstraction
based on them, will be first class, in keeping with the spirit of functional programming. Programs implemented in FRP languages
should of course be efficiently executable, but this has proven to be the main challenge in implementing FRP.

@@

A recent thesis~\cite{Czaplicki2012} described Elm, a stand-alone language for
reactivity. Elm provides combinators for manipulating discrete events, and
compiles to Javascript, making it useful for client-side web programming.
However, Elm does not provide a notion of switching or continuous time
behaviors, though an approximation is given using discrete-time events which are
actuated at repeated intervals specified during the event definition. The thesis
asserts that Arrowized FRP (signal-function FRP, Section~\ref{section:Background-signal_function_frp})
can be embedded in Elm, but provides little support for this assertion\footnote{
A form of Arrowized FRP employing applicative functors is presented, and
justified by the assertion that applicative functors are just arrows with the
input type parameter fixed. While this is true, it ignores the compositional
benefits of the general Arrow framework, and it is not clear that it provides
the same enforcement of causality.}

@@

An alternative approach to FRP was first proposed in work on
Fruit~\cite{Courtney2001-1}, a library for declarative specification of GUIs.
This library utilized the abstraction of Arrows~\cite{Hughes2000} to structure
{\em signal functions}. Arrows are abstract type constructors with input and
output type parameters, together with a set of routing combinators (see Figure~\ref{figure:arrow_combinators}).
Signal functions are the first-class abstraction in this approach. They
represent time-dependent and reactive transformers of events and signals, which
are themselves not first class, since such values cannot be directly manipulated
by the programmer. This approach has two motivations: it increases modularity
since both the input and output of signal functions may be transformed (as
opposed to signals or events which may only be transformed in their output) and
it avoids a large class of time and space leaks which have emerged when
implementing FRP with first-class signals and events.

@@

Similarly to FrTime, the netwire~\cite{Soylemez} library eschews dynamic
switching, in this case in favor of {\em signal inhibition}. Netwire is written
as an arrow transformer, which, together with the Kliesli arrow instance for the
IO monad\footnote{Any monad forms a Kliesli arrow.}, permits it to lift IO
actions as sources and sinks at arbitrary points in a signal function network.
Signal inhibition is accomplished by making the output of signal functions a
monoid, and then combining the outputs of signal functions. An inhibited signal
function will produce the monoid's zero as an output. Primitives have defined
inhibition behavior, and composed signal functions inhibit if their outputs
combine to the monoid's zero.

@@

A recent PhD thesis~\cite{Sculthorpe2011} introduced N-Ary FRP, a technique for
typing Arrowized FRP systems using dependent types. The bulk of the work
consisted in using the dependent type system to prove the correctness of the FRP
system discussed. This work introduced the typing construct of signal vectors,
which permit the distinction of signal and event types at the level of the FRP
system, instead of making events merely a special type of signal.

>>>>

The key abstraction of functional programming is the function, which takes an
input and produces an output. Functions may produce functions as output and/or
take functions as input. We say that functions are {\em first-class values} in a
functional programming language.

@@

The goal of FRP is to provide compositional and high-level abstractions for
creating reactive programs. The key abstractions of FRP are behaviors or
signals\footnote{Behaviors are generally referred to as {\em signals} in
signal-function literature. This is unfortunate, since a {\em signal function}
may operate on events, signals, or some combination of the two.}, which are
time-dependent values defined at every point in continuous time, and events,
which are values defined at countably many points in time. An FRP system will
provide combinators to manipulate events and signals and to react to events by
replacing a portion of the running program in response to an event. Behaviors
and events, or some abstraction based on them, will be first class, in keeping
with the spirit of functional programming. Programs implemented in FRP languages
should of course be efficiently executable, but this has proven to be the main
challenge in implementing FRP.

@@

A recent thesis~\cite{Czaplicki2012} described Elm, a stand-alone language for
reactivity. Elm provides combinators for manipulating discrete events, and
compiles to JavaScript, making it useful for client-side web programming.
However, Elm does not provide a notion of switching or continuous time
behaviors, though an approximation is given using discrete-time events which are
actuated at repeated intervals specified during the event definition. The thesis
asserts that Arrowized FRP (signal-function FRP, Section~\ref{section:Background-signal_function_frp})
can be embedded in Elm, but provides little support for this assertion\footnote{
A form of Arrowized FRP employing applicative functors is presented, and
justified by the assertion that applicative functors are just arrows with the
input type parameter fixed. While this is true, it ignores the compositional
benefits of the general Arrow framework, and it is not clear that it provides
the same enforcement of causality.}

@@

An alternative approach to FRP was first proposed in work on
Fruit~\cite{Courtney2001-1}, a library for declarative specification of GUIs.
This library utilized the abstraction of Arrows~\cite{Hughes2000} to structure
{\em signal functions}. Arrows are abstract type constructors with input and
output type parameters, together with a set of routing combinators (see Figure~\ref{figure:arrow_combinators}).
Signal functions form an arrow. Signal functions are the first-class abstraction
in this approach. They represent time-dependent and reactive transformers of
events and signals, which are themselves not first class, since such values
cannot be directly manipulated by the programmer. This approach has two
motivations: it increases modularity since both the input and output of signal
functions may be transformed (as opposed to signals or events which may only be
transformed in their output) and it avoids a large class of time and space leaks
which have emerged when implementing FRP with first-class signals and events.

@@

Similarly to FrTime, the netwire~\cite{Soylemez} library eschews dynamic
switching, in this case in favor of {\em signal inhibition}. Netwire is written
as an arrow transformer, and permits the lifting of IO actions as sources and
sinks at arbitrary points in a signal function network. Signal inhibition is
accomplished by making the output of signal functions a monoid, and then
combining the outputs of signal functions. An inhibited signal function will
produce the monoid's zero as an output. Primitives have defined inhibition
behavior, and composed signal functions inhibit if their outputs combine to the
monoid's zero.

@@

A recent PhD thesis~\cite{Sculthorpe2011} introduced N-Ary FRP, a technique for
typing Arrowized FRP systems using dependent types. The bulk of the work
consisted in using the dependent type system to prove the correctness of the FRP
system discussed. This work introduced a signal vectors, a typing construct
which permits the distinction of signal and event types at the level of the FRP
system, instead of making events merely a special type of signal.

### 2012-11-05 Docs/Thesis/comparison.tex
Thesis edits

The application of choice is a learning switch controller for OpenFlow~\cite{OpenflowSpec}
switches. this application must handle input events triggered by switches
signalling that they have received a packet for which they have no handling
rules, and produce output events which install handling rules for dealing with
these packets.

@@

The controllers were benchmarked using the {\tt cbench} program~\cite{cbench}. The
amount to time to pause between time and signal updates was passed as a parameter to
both programs. The {\tt cbench} benchmark operates by sending a simulated packet in
event, waiting for a rule response, and then repeating. The output of the benchmark
is the number of times this occurs in a second. This is thus a very good test of
event latency in a reactive system.

>>>>

The application of choice is a learning switch controller for OpenFlow~\cite{OpenflowSpec}
switches. This application must handle input events triggered by switches
signalling that they have received a packet for which they have no handling
rules, and produce output events which install handling rules for dealing with
these packets.

@@

The controllers were benchmarked using the {\tt cbench} program~\cite{cbench}. The
amount to time to pause between time and signal updates (the inverse of the sampling rate for signals)
was passed as a parameter to both programs. The {\tt cbench} benchmark operates by sending a simulated packet in
event, waiting for a rule response, and then repeating. The output of the benchmark
is the number of times this occurs in a second. This is thus a very good test of
event latency in a reactive system.

### 2012-11-05 Docs/Thesis/conclusions-and-further-work.tex
Thesis edits

Further, the model of events used by TimeFlies subverts problematic semantic
questions about the evaluation of events in an FRP system. By using the N-Ary
FRP type model and separating the evaluation of events from the time steps used
for signals, TimeFlies fully supports non-deterministic merging of events,
and provides a semantic guarantee that events are not ''lost'' during evaluation.

@@

In order to improve the performance of FRP yet further, it may be productive to attempt
to introduce parallel evaluation into FRP, taking advantage of the functional purity in
the implementation of the signal function combinators. This may involve, for instance,
evaluating several time-steps at once in a data-parallel manner,task parallelism
between different branches of a signal function, or speculative evaluation of switch
combinators.

>>>>

Further, the model of events used by TimeFlies subverts problematic semantic
questions about the evaluation of events in an FRP system. By using the N-Ary
FRP type model and separating the evaluation of events from the time steps used
for signals, TimeFlies fully supports non-deterministic merging of events,
and provides a semantic guarantee that events are not ``lost'' during evaluation.

@@

In order to improve the performance of FRP yet further, it may be productive to attempt
to introduce parallel evaluation into FRP, taking advantage of the functional purity in
the implementation of the signal function combinators. This may involve, for instance,
evaluating several time-steps at once in a data-parallel manner, task parallelism
between different branches of a signal function, or speculative evaluation of switch
combinators.

### 2012-11-05 Docs/Thesis/implementation-evalinterface.tex
Thesis edits

This property will be useful because we wish our evaluation interface to be
part of a monad stack, the base of which is Haskell's IO monad. By formulating
the evaluation interface as a monad transformer, we need only define the
operations relevant to the evaluation of signal functions, and we can depend
on the constructs of the IO monad to interact with whatever inputs and outputs
are necessary. In some cases, we may not wish to use the IO monad at all (e.g
for testing or simulation). In this case, we can parameterize over another
monad, such as the Identity monad (which has no special operations and whose
context is just the value), or the State monad (which maintains an implicit 
state accessible by {\tt get} and {\tt put} operations).

@@

The evaluation interface is exported as follows:

\begin{code}
data SFEvalState svIn m
data SFEvalT svIn m
data HandlerSet sv m
data EventOccurrence sv
data SignalDelta sv

sd      :: a -> SignalDelta (SVSignal a)
sdLeft  :: SignalDelta svl -> SignalDelta (SVAppend svl svr)
sdRight :: SignalDelta svr -> SignalDelta (SVAppend svl svr)
sdBoth  :: SignalDelta svl -> SignalDelta svr 
        -> SignalDelta (SVAppend svl svr)

eo      :: a -> EventOccurrence (SVEvent a)
eoLeft  :: EventOccurrence svl -> EventOccurrence (SVAppend svl svr)
eoRight :: EventOccurrence svr -> EventOccurrence (SVAppend svl svr)

sdHS    :: (a -> m ()) -> HandlerSet m (SVSignal a)
eoHS    :: (a -> m ()) -> HandlerSet m (SVEvent a)
hsLeft  :: HandlerSet m svl -> HandlerSet m (SVAppend svl svr)
hsRight :: HandlerSet m svr -> HandlerSet m (SVAppend svl svr)
hsBoth  :: HandlerSet m svl -> HandlerSet m svr 
        -> HandlerSet m (SVAppend svl svr)

initSFEval :: HandlerSet svOut m 
           -> SF NonInitialized svIn svOut 
           -> SFEvalState svIn m

runSFEvalT :: SFEvalT svIn m a -> SFEvalState svIn m -> m a

push   :: EventOccurrence svIn -> SFEvalT svIn m ()
update :: SignalDelta svIn -> SFEvalT svIn m ()
sample :: Double -> SFEvalT svIn m ()
\end{code}

@@

>>>>

This property will be useful because we want our evaluation interface to be
part of a monad stack, the base of which is Haskell's IO monad. By formulating
the evaluation interface as a monad transformer, we need only define the
operations relevant to the evaluation of signal functions, and we can depend
on the constructs of the IO monad to interact with whatever inputs and outputs
are necessary. In some cases, we may not wish to use the IO monad at all (e.g
for testing or simulation). In this case, we can parameterize over another
monad, such as the Identity monad (which has no special operations and whose
context is just the value), or the State monad (which maintains an implicit 
state accessible by {\tt get} and {\tt put} operations).

@@

The evaluation interface is exported as shown in
Section~\ref{section:System_Design_and_Interface-Evaluator}.

@@

The {\tt update} action updates the current input signal delta, which will be
sampled at the next {\tt step} action. It has no immediately observable effect.
It simply updates the value of one signal in the input signal vector, a delta
of which is stored in the {\tt SFEvalState} record.

The {\tt step} action takes a time delta, and calls the signal continuation of
the stored signal function. It actuates the outputs using the stored handlers,
replaces the stored delta with an empty delta, and stores the resulting new
signal function in the state.

### 2012-11-05 Docs/Thesis/implementation-signalfunctions.tex
Thesis edits

In order to create these concrete representations, we employ the Haskell type
system extension known as Generalized Algebraic Datatypes, which we divert
briefly to explain before exploring its use in the implementation.

@@

In Chapter~\ref{chapter:System_Design_and_Interface} we present signal vectors
as a set of types. In order to be completely principled, we should isolate these
types into their own {\tt kind} (a sort of type of types), however, the Haskell
extension for this was far from stable at the time this system was created.

@@

To get our feet wet, we create a representation which carries a value for
every {\tt SVSignal} leaf of a signal vector. In order to do this, we restrict
each of our constructors to a single signal vector type. This ensures that the
only way to represent a sample leaf is with the {\tt SVSample} constructor,
which carries a value of the appropriate type. The datatype is shown in
Figure~\ref{figure:signal_sample_datatype}.

@@

So we now have a representation for the signal components of a signal vector.
What about the event components? We want to represent event occurrences, each
of which will correspond to at most one event in the vector. So a different
representation is called for. In this case, there will be only three
constructors. One constructor will represent an event leaf, and the other will
represent a single value on the left or right side of the node ({\tt SVAppend}),
ignoring all of the type structure on the other side. This representation
describes a path from the root of the signal vector, terminating at an event
leaf with a value.

@@

Another strategy is notification, in which the program exposes an interface
which the source of updated information may invoke. This is a repeated entry
point to the program, which causes the program to perform the same tasks
listed above, namely, evaluate the updated information, update state, and
produce output. The strategy of notification as opposed to repeated checking is
the essence of push-based evaluation.

@@

The simplest combinator is the {\tt identity} signal function. This signal
function simply passes all of its inputs along as outputs. The initialization
function simply passes along the received sample and outputs the initialized
version of the signal function. The initialized version of the input is similar,
but is self-referential. It replaces itself with itself after every signal
function or event. The implementation is shown in Figure~\ref{figure:identity_implementation}.

@@

In order for our primitive signal functions to be useful, we need a means of
composing them. Serial composition creates one signal function from two, by
using the output of one as the input of the other. The serial composition
combinator is styled {\tt (>>>)}. The implementation of this operator is one
place where the advantage of responding to events independently from signal
samples becomes clear. The implementation is shown in Figure~\ref{figure:serial_composition_implementation}.

@@

The implementation of the non-initialized version is simple. It takes two
non-initialized signal functions, and returns the sample output of applying the
second to the sample output of the first, along with an initialized signal
function constructed by applying the initialized version of the combinator to
the initialized signal functions given by the inputs.

The initialized version constructs two continuations. The first receives a time
and signal update input, and applies the signal continuation of the first
signal function to it. The output is fed, along with the time update, into the
second signal function, first the signal update, and then events. The output
of the second, along with the accumulated updated signal functions with the
initialized serial composition recursively applied, is returned as output.

The event continuation is similar but simpler, as there is no need to deal
with signal updates.

The switch combinator is the means of introducing reactivity into a signal
function. This combinator allows a signal function to replace itself by
producing an event occurrence. The switch combinator stores the full sample
for its input vector (which is identical to the input vector of the supplied
signal function) to initialize the new signal function. This also demands that
it continue wrapping the signal function until the next sampling interval, so
that it can actuate the output of initialization as a replacement signal. This
has some performance implications, which are discussed in
Chapter~\ref{chapter:Evaluation_and_Comparisons}. The implementation is shown in
Figure~\ref{figure:switch_implementation}.

\begin{figure}
\begin{code}
switch :: (svIn :~> (svOut :^: SVEvent (svIn :~> svOut))) -> svIn :~> svOut
switch (SF sigSampleF) =
  SF (\sigSample -> let (sigSampleSF, sf) = sigSampleF sigSample
                        (sigSampleOut, _) = splitSample sigSampleSF
                    in (sigSampleOut, switchInit sigSample sf))

-- Pre-occurrence
switchInit ::    SVSample svIn
              -> SF Initialized svIn 
                    (SVAppend svOut 
                     (SVEvent (SF NonInitialized svIn svOut)))
              -> SF Initialized svIn svOut
switchInit inputSample sf@(SFInit timeCont changeCont) =
  SFInit (\dt sigDelta ->
            let (sigDeltaSF, occsSF, newSF) = timeCont dt sigDelta
                newInputSample = updateSample inputSample sigDelta
                (sigDeltaOut, _) = splitDelta sigDeltaSF
                (occsOut, occsSwitch) = splitOccurrences occsSF
                (outputDelta, nextSF) =
                  maybe (sigDeltaOut, switchInit newInputSample newSF)
                    (\(SF sigSampleF) ->
                       let (outputSample, switchSF) =
                             sigSampleF newInputSample
                           outputDelta = sampleDelta outputSample
                       in (outputDelta, switchSF))
                  $ occurrenceListToMaybe occsSwitch
            in (outputDelta, occsOut, nextSF))
         (\evtOcc ->
            let (occsSF, newSF) = changeCont evtOcc
                (occsOut, occsSwitch) = splitOccurrences occsSF
                nextSF = maybe (switchInit inputSample newSF) 
                               (\(SF sigSampleF) ->
                                  let (sampleOut, switchSF) =
                                        sigSampleF inputSample
                                  in switchWait sampleOut switchSF)
                         $ occurrenceListToMaybe occsSwitch
            in (occsOut, nextSF))

switchWait ::    SVSample svOut
              -> SF Initialized svIn svOut
              -> SF Initialized svIn svOut
switchWait outputSample sf@(SFInit timeCont changeCont) =
  SFInit (\dt sigDelta -> let (sigDeltaOut, outOccs, newSF) =
                                timeCont dt sigDelta
                          in (sampleDelta $
                                updateSample outputSample sigDeltaOut,
                              outOccs,
                              newSF))
         \evtOcc -> let (outOccs, newSF) = changeCont evtOcc
                    in (outOccs, switchWait outputSample newSF))
\end{code} 
\hrule
\caption{Implementation of reactivity.}
\label{figure:switch_implementation}
\end{figure}

@@

\end{code}
\hrule
\caption{Swap routing combinator implementation.}
\label{figure:swap_implementation}
\end{figure}

The {\tt first} and {\tt second} combinators are not quite as simple, as they
must transform a provided signal function. For signal changes, they must split
the set of changes into those which will be passed to the signal function and
those which will be simply passed along to the output, and then recombine them
on the other side. For event occurrences, the occurrence must be pattern-matched
to determine whether to call the event continuation from the provided signal
function or passed through, and output event occurrences must have the suitable
routing constructor re-applied. In any case, when a continuation has been
applied, the combinator must be recursively applied to the new signal function.
The implementation of {\tt first} is shown in Figure~\ref{figure:first_implementation}.
The implementation of {\tt second} is the obvious reversal of this
implementation.

\begin{figure}
\begin{code}
first :: (svIn :~> svOut) -> (svIn :^: sv) :~> (svOut :^: sv)
first (SF sigSampleF) =
  SF (\sigSample -> let (leftSample, rightSample) = splitSample sigSample
                        (leftSampleOut, sf) = sigSampleF leftSample
                    in (combineSamples leftSampleOut rightSample,
                        firstInit sf))

firstInit ::    SF Initialized svIn svOut
             -> SF Initialized (SVAppend svIn sv) (SVAppend svOut sv)
firstInit (SFInit timeCont inputCont) = 
  let firstInitSF =
    SFInit (\dt sigDelta ->
               let (input, rightOutput) = splitDelta sigDelta
                   (sigDeltaOutput, evtOutput, sf1New) = timeCont dt input
               in (combineDeltas sigDeltaOutput rightOutput,
                   map occLeft evtOutput,
                   firstInit sf1New))
           (\evtOcc ->
              case chooseOccurrence evtOcc of
                Left  lChange -> let (changes, sf) = inputCont lChange
                                 in (map occLeft changes, firstInit sf)
                Right rChange -> ([occRight rChange], firstInitSF))
  in firstInitSF
\end{code}
\hrule
\caption{Implemenation of {\tt first} combinator.}
\label{figure:first_implementation}
\end{figure}

The looping feedback combinator is particularly tricky. As it is currently
implemented, the initial sample for the right side of the input signal vector to
the supplied function is the right side of the output sample. This is acceptable,
given Haskell's non-strict evaluation strategy, but it is necessary that the
right side of the signal function's output not be immediately dependent on its
input. The feedback combinator makes use of Haskell's lazy evaluation to
feed events back into the combinator, and stores signal updates until the next
sample. Signal samples are thus automatically decoupled after initialization.
The implementation, which makes extensive use of the recursive nature of
Haskell's {\tt let} construct, is shown in Figure~\ref{figure:loop_implementation}.

\begin{figure}
\begin{code}
loop :: (svIn :^: svLoop) :~> (svOut :^: svLoop) -> svIn :~> svOut
loop (SF sigSampleF) =
  SF (\sigSample -> 
        let (sigSampleOut, sfInit) =
              sigSampleF (combineSamples sigSample sigSampleOutRight)
            (sigSampleOutLeft, sigSampleOutRight) =
              splitSample sigSampleOut
        in (sigSampleOutLeft, loopInit deltaNothing sfInit))

loopInit ::    SVDelta svLoop
            -> SF Initialized (SVAppend svIn svLoop) (SVAppend svOut svLoop)
            -> SF Initialized svIn svOut
loopInit loopDelta (SFInit timeCont changeCont) =
  SFInit (\dt sigDelta ->
            let (sigDeltaOut, occsOut, newSF) =
                  timeCont dt (combineDeltas sigDelta loopDelta)
                (occsOutput, loopOccs) = splitOccurrences occsOut
                (occsOutput', loopOccs') = splitOccurrences foldOccs
                (foldOccs, newSF') =
                  applySF newSF (map occRight (loopOccs ++ loopOccs'))
                (outputDelta, newLoopDelta) = splitDelta sigDeltaOut
            in (outputDelta,
                occsOutput ++ occsOutput',
                loopInit newLoopDelta newSF'))
         (\change -> let (occsOut, newSF) = changeCont $ occLeft change
                         (occsOutput, loopOccs) =
                           splitOccurrences occsOut
                         (occsOutput', loopOccs') =
                           splitOccurrences foldOccs
                         (foldOccs, newSF') =
                           applySF newSF
                             (map occRight (loopOccs ++ loopOccs'))
                     in (occsOutput ++ occsOutput',
                         loopInit loopDelta newSF'))
\end{code}
\hrule
\caption{Implementation of feedback.}
\label{figure:loop_implementation}
\end{figure}

@@

The implementation strategy leaves room for optimizations. In particular, an
additional constructor for time-independent signal functions would allow
portions of a signal function to forgo evaluation during time steps unless they
had signal updates. Optimizations in the style of Yampa, observed by keeping
an updated for the signal function and pattern-matching on it when switching,
might further improve performance. Nevertheless, this implementation performs
quite well as it currently exists, as we demonstrate in Chapter~\ref{chapter:Evaluation_and_Comparisons}.

>>>>

@@

In Chapter~\ref{chapter:System_Design_and_Interface} we presented signal vectors
as a set of types. In order to be completely principled, we should isolate these
types into their own {\em kind} (a sort of type of types); however, the Haskell
extension for this was far from stable at the time this system was created.

@@

The first thing to represent is {\em samples}, which are sets of values for
the signal components of a signal vector. Therefore, we create a representation
which carries a value for every {\tt SVSignal} leaf of a signal vector. In order
to do this, we restrict each of our constructors to a single signal vector type.
So there are three leaf constructors: {\tt SVSample}, which carries a value; and
{\tt SVSampleEvent} and {\tt SVSampleEmpty}, which do not. This ensures that the
only way to represent a sample leaf is with the {\tt SVSample} constructor,
which carries a value of the appropriate type. The datatype is shown in
Figure~\ref{figure:signal_sample_datatype}.

@@

What about the event components? We want to represent event occurrences,
each of which will correspond to at most one event in the vector. So a different
representation is called for. In this case, there will be only three
constructors. One constructor will represent an event leaf, and the other will
represent a single value on the left or right side of the node ({\tt SVAppend}),
ignoring all of the type structure on the other side. This representation
describes a path from the root of the signal vector, terminating at an event
leaf with a value.

@@

Another approach is notification, in which the program exposes an interface
which the source of updated information may invoke. This is a repeated entry
point to the program, which causes the program to perform the same tasks
listed above, namely, evaluate the updated information, update state, and
produce output. The strategy of notification as opposed to repeated checking is
the essence of push-based evaluation.

@@

As an example of the implementation of combinators, we show the implementation
of the {\tt identity} signal function in Figure~\ref{figure:identity_implementation}.
This signal function simply passes all of its inputs along as outputs. The
initialization function simply passes along the received sample and outputs the
initialized version of the signal function. The initialized version of the input
is similar, but is self-referential. It outputs itself as its replacement. This
is standard for simple and routing combinators which are not reactive, and
simply move samples and event occurrences around.

@@

In order for our primitive signal functions to be useful, we need a means of
composing them. Serial composition creates one signal function from two, by
using the output of one as the input of the other. The serial composition
combinator is styled {\tt (>>>)}. The implementation of this operator is one
place where the advantage of responding to events independently from signal
samples becomes clear.

This is the only primitive combinator which takes two signal functions, and
thus, it is the only way to combine signal functions. Parallel, branching, and
joining composition can be achieved by modifying signal functions with the
{\tt first} and {\tt second} combinators and composing them with the
routing and joining combinators.

Combinators which take one or more signal functions as input must recursively
apply themselves, as is shown in the implementation of serial composition
(Figure~\ref{figure:serial_composition_implementation}). They must also
handle initialization, retaining the initialized signal functions and passing
them to the initialized version of the combinator.

@@

The switch combinator is the means of introducing reactivity into a signal
function. This combinator allows a signal function to replace itself by
producing an event occurrence. The combinator wraps a signal function, and 
observes an event on the right side of the output signal vector. At the first
occurrence of the event, the signal function carried by the occurrence replaces
the signal function.

The switch combinator stores the input sample provided during initialization,
and updates it with the input signal updates. When the wrapped signal function
produces an occurrence carrying a new signal function, that signal function is
initialized with the stored input sample. It is then "wrapped" by another
function which closes over its output sample, and outputs the sample as a signal
update as the next time step. After this, it acts as the new signal function.
This wrapping has some performance implications, which are discussed

This combinator checks the outputs of the wrapped
signal function for an event occurrence from which an uninitialized signal
function is extracted. The switch combinator stores the full sample
for its input vector (which is identical to the input vector of the supplied
signal function) to initialize the new signal function. This also demands that
it add a wrapper to the new signal function which waits for the next sampling
interval and actuates the sample output at initialization as an output set
of changes to the signal. This has some performance implications, which are
discussed in Chapter~\ref{chapter:Evaluation_and_Comparisons}.

@@

\end{code}
\hrule
\caption{Implementation of the {\tt swap} routing combinator.}
\label{figure:swap_implementation}
\end{figure}

The {\tt first} and {\tt second} combinators are similar to serial composition,
but they transform only one signal function. For signal changes, they must split
the set of input changes into those which will be passed to the signal function
and those which will be simply passed along to the output, and then recombine
them on the other side. For event occurrences, the occurrence must be
pattern-matched to determine whether to call the event continuation from the
provided signal function or passed through, and output event occurrences must
have the suitable routing constructor re-applied. In any case, when a
continuation has been applied, the combinator must be recursively applied to the
new signal function.

The looping feedback combinator is particularly tricky. As it is currently
implemented, the initial sample for the right side of the input signal vector to
the supplied function is the right side of the output sample. This is acceptable,
given Haskell's non-strict evaluation strategy, but it is necessary that the
right side of the signal function's output not be immediately dependent on its
input. The feedback combinator makes use of Haskell's lazy evaluation to
feed events back into the combinator, and stores signal updates until the next
sample. Signal samples are thus automatically decoupled after initialization.
The implementation makes use of the recursive nature of the {\tt let} construct
in Haskell, and the non-strict evaluation of Haskell, to implement feedback.

@@

The implementation strategy leaves room for optimizations. In particular, an
additional constructor for time-independent signal functions would allow
portions of a signal function to forgo evaluation during time steps unless they
had signal updates. Optimizations in the style of Yampa, observed by keeping
an updated AST for the signal function and pattern-matching on it when switching,
might further improve performance. In particular, collapsing nested or
serially-composed versions of the {\tt switchWait} step when switching would
remove at least some of the observed dependence of performance on sampling rate.
Nevertheless, this implementation performs quite well as it currently exists, as
we demonstrate in Chapter~\ref{chapter:Evaluation_and_Comparisons}.

### 2012-11-05 Docs/Thesis/implementation.tex
Thesis edits

Having established the design and semantics of the reactive system, I present a purely 
functional implementation. For this implementation, I will make significant use
of an extension to the Haskell type system. This extension, called
GADTs~\cite{Cheney2003,Xi2003}, provides a mechanism for expressing abstractions
which are not easily expressible in a standard Hindley-Milner type system, while
maintaining the type soundness of Haskell. In particular, GADTs provide the
power we require to take advantage of the signal vector representation of signal
function inputs and outputs.

>>>>

Having established the design and semantics of the reactive system, I present a purely 
functional implementation. For this implementation, I will make significant use
of an extension to the Haskell type system. This extension, called
Generalized Algebraic Datatypes (GADTs)~\cite{Cheney2003,Xi2003}, provides a mechanism
for expressing abstractions which are not easily expressible in a standard
Hindley-Milner type system, while maintaining the type soundness of Haskell.
In particular, GADTs provide the power we require to take advantage of the
signal vector representation of signal function inputs and outputs.
